Performing graph-based clustering with rOpenScPCA

### 2024-11-15 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
Finish notebook rearrangement

---
title: "Performing and evaluating graph-based clustering with rOpenScPCA"
date: "`r Sys.Date()`"
author: "Data Lab"
params:
  seed: 2024
output: 
  html_notebook: 
    toc: yes
    toc_float: yes
    df_print: paged
---

@@

This notebook provides examples of how to use the `rOpenScPCA` package to:

- Perform clustering
- Evaluate clustering results using several metrics, including:
  - Silhouette width
  - Neighborhood purity
  - Cluster stability, as measured with the Adjusted Rand Index
- Save clustering results

While this notebook does demonstrate how to use individual functions that evaluate a single clustering result, a full clustering evaluation would entail comparison these metrics across different clusterings from different parameterizations.

To demonstrate these tasks, this notebook will use the sample `SCPCS000001` from project `SCPCP000001`, which is assumed present in the `OpenScPCA-analysis/data/current/SCPCP000001` directory.
Please [see this documentation](https://openscpca.readthedocs.io/en/latest/getting-started/accessing-resources/getting-access-to-data/) for more information about obtaining ScPCA data.

@@

suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(Seurat)
  library(dplyr)
  library(ggplot2)
})

# Set ggplot theme
theme_set(theme_bw())
```

@@

### Read in data

To begin, we'll read in the `SingleCellExperiment` (SCE) object.

@@

We'll also extract its PCA matrix, which we'll use to demonstrate how to calculate and evaluate clusters.

```{r extract pca data}
# Extract the PCA matrix
pca_matrix <- reducedDim(sce, "PCA")
```

If you are working with a Seurat object, you can extract the PCA matrix like this:

```{r convert to seurat}
# First, convert the SCE to a Seurat object for demonstration

# Here, the PCA matrix has the same name as a
```

@@

* The `algorithm` can be one of:
  * `louvain`, `walktrap`, or `leiden`
* The `weighting` can be one of:
  * `jaccard`, `rank`, or `number`
* The nearest neighbors parameter can be customized with the `nn` argument
* The resolution parameter can be customized with the `resolution` argument
  * This parameter is only used by Louvain and Leiden algorithms
* If the Leiden algorithm is used, its default objective function parameter will be `CPM`, but you can also set  `objective_function = "modularity` instead.
* You can provide additional parameters as a list to the `cluster_args` argument.
  * Please refer to the [`igraph` documentation](https://igraph.org/r/html/latest) to learn more about what additional parameters can be provided to each clustering algorithm.
  * Note that `cluster_args` only accepts single-length arguments (no vectors or lists).

@@

## Evaluate clustering results

@@

```{r stability, warning=FALSE}
# calculate the stability of clusters
stability_results <- calculate_stability(
  sce,
  cluster_results_df
)

@@

```{r stability custom parameters}
# Calculate clusters with non-default parameters
cluster_df_leiden <- calculate_clusters(
  sce,
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15
)

# Now, pass in the same arguments customizing parameters here
stability_results_leiden <- calculate_stability(
  sce,
  cluster_df_leiden,
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15
)

@@

## Evaluating pre-calculated clusters

Even if you perform clustering with other tools, you can still use the `rOpenScPCA` functions to evaluate your clusters.
For example, if you are analyzing your data with a Seurat pipeline, you might also calculate clusters with Seurat, and you can evaluate those with `rOpenScPCA`.
In addition, ScPCA data already contains a column called `cluster` with results from an automated clustering.

In this section, we'll present how you can calculate the silhouette width and neighborhood purity from existing cluster assignments.

@@

First, we'll prepare a Seurat object and perform clustering for demonstration:

```{r process seurat object}
# Convert to a Seurat object
seurat_obj <- CreateSeuratObject(counts = counts(sce), assay = "RNA")

# Process the object with a Seurat pipeline to obtain clusters
seurat_obj <- seurat_obj |>
  SCTransform() |>
  RunPCA() |>
  FindNeighbors(dims = 1:30) |>
  FindClusters()

# Print resulting clusters
head(seurat_obj$seurat_clusters)
```

Next, we'll prepare inputs for `rOpenScPCA` functions:

@@

```{r prepare seurat input}
# Extract the PCA matrix
seurat_pca_matrix <- Embeddings(seurat_obj, reduction = "pca")

# Create a data frame
seurat_cluster_df <- data.frame(
  cell_id = names(seurat_obj$seurat_clusters),
  cluster = seurat_obj$seurat_clusters
)
```

@@

First, you should prepare a data frame two columns: `cell_id` containing cell barcodes, and `cluster` containing the cluster identities.

```{r}
scpca_cluster_df <- data.frame(
  cell_id = colnames(sce),
  cluster = colData(sce)$cluster
)

@@

Then, you can proceed to run evaluation functions using this data frame, and the PCA matrix from the object.

```{r}
# Calculate silhouette width
scpca_silhouette_df <- calculate_silhouette(pca_matrix, scpca_cluster_df)
scpca_silhouette_df

# Calculate neighborhood purity
scpca_purity_df <- calculate_purity(pca_matrix, scpca_cluster_df)
scpca_purity_df
```

When running `calculate_stability()`, it is important to specify the same parameters originally used to build the clusters:

```{r}
scpca_stability_df <- calculate_stability(
  pca_matrix,
  scpca_cluster_df,
  # provide ScPCA clustering parameters
  algorithm = "louvain",
  weighting = "jaccard",
  nn = 20,
  resolution = 1
)
scpca_stability_df
```

@@

Instead of a matrix, you can alternatively pass in an SCE or Seurat object that contains PCs.
For example:

```{r}
# Calculate clusters from an SCE object using default parameters
cluster_results_df <- calculate_clusters(sce)

# Calculate clusters from a Seurat object using default parameters
cluster_results_df <- calculate_clusters(seurat_obj)
```

`rOpenScPCA` assumes that the PCA matrix is named `PCA` in SCE objects, and `pca` in Seurat objects.
If the PCA matrix you want to use in the object has a different name, you can provide the argument `pc_name`.
For example:

```{r}
# First, we'll rename the PCA matrix for demonstration
reducedDimNames(sce) <- c("PCA_matrix", "UMAP")

# Calculate clusters from an SCE object using default parameters
cluster_results_df <- calculate_clusters(
  sce,
  pc_name = "PCA_matrix"
)
```

@@

Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv()`), or you can add the results into your SCE or Seurat object.
The subsequent examples will show you, as an example, how to save the `cluster_results_df` data frame to a given object.

@@

To avoid ambiguity between the existing and new clustering results, you may want to rename incoming columns.
We'll therefore begin by adding the prefix `ropenscpca_` to all column names in the `cluster_results_df` data frame before adding them to a given object.

@@

# Third, add columns as desired using the barcodes and cell_id columns to join
sce_coldata <- sce_coldata |>
  dplyr::left_join(
    cluster_results_renamed_df,
    by = c("barcodes" = "ropenscpca_cell_id")
  )

# Finally, add the modified colData back into the SCE
# Use the row.names argument to ensure rownames (barcodes) are retained; don't forget this step!
colData(sce) <- DataFrame(
  sce_coldata,
  row.names = sce_coldata$barcodes
)

# Print updated colData slot
colData(sce)
```

>>>>

---
title: "Performing graph-based clustering with rOpenScPCA"
date: "`r Sys.Date()`"
author: "Data Lab"
params:
  seed: 2024
output: 
  html_notebook: 
    toc: yes
    toc_float: yes
    df_print: paged
---

@@

This notebook provides examples of how to use functions in `rOpenScPCA` that:

* Perform clustering
* Calculate QC metrics on clustering including:
  * Silhouette width
  * Neighborhood purity
  * Cluster stability, as measured with the Adjusted Rand Index
* Calculate QC metrics obtained with other tools, such as `Seurat`
* Save clustering results to an SCE or `Seurat`

While this notebook demonstrates how to use individual functions that calculate helpful metrics for evaluating clustering results, a full evaluation would compare these metrics across different clusterings from different parameterizations.

This notebook will use the sample `SCPCS000001` from project `SCPCP000001`, which is assumed present in the `OpenScPCA-analysis/data/current/SCPCP000001` directory, for all examples.
Please [see this documentation](https://openscpca.readthedocs.io/en/latest/getting-started/accessing-resources/getting-access-to-data/) for more information about obtaining ScPCA data.

@@

suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(Seurat)
  library(dplyr)
})
```

@@

## Read in and prepare data

To begin, we'll read in the `SingleCellExperiment` (SCE) object.
We'll also establish a corresponding processed Seurat object from its raw counts that we'll use for some examples.

@@

```{r create process seurat, message = FALSE}
# Convert to a Seurat object
seurat_obj <- CreateSeuratObject(counts = counts(sce), assay = "RNA")

# Process the object with a Seurat pipeline to obtain clusters
seurat_obj <- seurat_obj |>
  SCTransform() |>
  RunPCA() |>
  FindNeighbors() |>
  FindClusters()
```

We'll also extract its PCA matrix, which we'll use to demonstrate how to calculate and evaluate clusters.
We'll show how to do this for both SCE and Seurat objects:

```{r extract pca data}
# Extract the PCA matrix from an SCE object
# Moving forward, we'll use this version for most examples
pca_matrix <- reducedDim(sce, "PCA")

# Extract the PCA matrix from a Seurat object
seurat_pca_matrix <- Embeddings(seurat_obj, reduction = "pca")
```

@@

* The `algorithm` can be one of:
  * `louvain`, `walktrap`, or `leiden`
* The `weighting` can be one of:
  * `jaccard`, `rank`, or `number`
* The nearest neighbors parameter can be customized with the `nn` argument
* The resolution parameter can be customized with the `resolution` argument
  * This parameter is only used by Louvain and Leiden algorithms
* If the Leiden algorithm is used, its default objective function parameter will be `CPM`, but you can also set  `objective_function = "modularity"` instead.
* You can provide additional parameters as a list to the `cluster_args` argument.
  * Please refer to the [`igraph` documentation](https://igraph.org/r/html/latest) to learn more about what additional parameters can be provided to each clustering algorithm.
  * Note that `cluster_args` only accepts single-length arguments (no vectors or lists).

@@

## Calculate QC metrics on clusters

@@

```{r stability, warning=FALSE}
# calculate the stability of clusters
stability_results <- calculate_stability(
  pca_matrix,
  cluster_results_df
)

@@

```{r stability custom parameters}
# Calculate clusters with non-default parameters
cluster_df_leiden <- calculate_clusters(
  pca_matrix,
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15
)

# Now, pass in the same arguments customizing parameters here
stability_results_leiden <- calculate_stability(
  pca_matrix,
  cluster_df_leiden,
  algorithm = "leiden",
  resolution = 0.5,
  nn = 15
)

@@

##  Calculate QC metrics on existing clusters

Even if you perform clustering with other tools, you can still use the `rOpenScPCA` functions to evaluate your clusters.

In this section, we'll present examples of how you can calculate the silhouette width, neighborhood purity, and cluster stability from existing cluster assignments.

@@

If you are analyzing your data with a Seurat pipeline that includes calculating clusters, you can use `rOpenScPCA` to evaluate them.
We previously created a Seurat object called `seurat_obj` with clusters:

```{r show seurat clusters}
head(seurat_obj$seurat_clusters)
```

To calculate QC metrics on these clusters, we'll need two pieces of information:

@@

We'll create those here:

```{r prepare seurat input}
# Extract the PCA matrix from the Seurat object
seurat_pca_matrix <- Embeddings(seurat_obj, reduction = "pca")

# Create a data frame for input
seurat_cluster_df <- data.frame(
  cell_id = names(seurat_obj$seurat_clusters),
  cluster = seurat_obj$seurat_clusters
)
```

@@

ScPCA cell metadata already contains a column called `cluster` with results from an automated clustering. 
These clusters were calculated using `bluster`, the same tool that `rOpenScPCA` uses, with the following parameters:

* The Louvain algorithm
* Jaccard weighting
* 20 nearest neighbors

In this example, we'll show how to use the cluster evaluation functions on these clusters.

To begin, we'll prepare a data frame with two columns: `cell_id` containing cell barcodes, and `cluster` containing the cluster identities.

```{r prepare scpca data frame}
scpca_cluster_df <- data.frame(
  cell_id = colnames(sce),
  cluster = colData(sce)$cluster
)

@@

We can run evaluation functions using this data frame and the PCA matrix from the object.

```{r scpca silhouette}
# Calculate silhouette width
scpca_silhouette_df <- calculate_silhouette(pca_matrix, scpca_cluster_df)
scpca_silhouette_df
```

```{r scpca purity}
# Calculate neighborhood purity
scpca_purity_df <- calculate_purity(pca_matrix, scpca_cluster_df)
scpca_purity_df
```

When running `calculate_stability()`, we'll specify the same parameters originally used to build the clusters.
We only recommend evaluating clusters with `calculate_stability()` if you know the original parameters used.

```{r scpca stability}
scpca_stability_df <- calculate_stability(
  pca_matrix,
  scpca_cluster_df,
  # provide ScPCA clustering parameters
  algorithm = "louvain",
  weighting = "jaccard",
  nn = 20
)
scpca_stability_df
```

@@

Instead of a matrix, you can alternatively pass in an SCE or Seurat object that contains a matrix.
We show examples of this below with `rOpenScPCA::calculate_clusters()`, but this will also work for any of the evaluation functions as well:

```{r run on sce}
# Calculate clusters from an SCE object using default parameters
cluster_results_df <- calculate_clusters(sce)
cluster_results_df
```

```{r run on seurat}
# Calculate clusters from a Seurat object using default parameters
cluster_results_df <- calculate_clusters(seurat_obj)
cluster_results_df
```

`rOpenScPCA` assumes that the PCA matrix is named `PCA` in SCE objects, and `pca` in Seurat objects (as they were in the above examples).
If the PCA matrix you want to use in the object has a different name, you can provide the argument `pc_name`.
We can see this below with an SCE object, for example:

```{r}
# First, we'll rename the PCA matrix for demonstration
reducedDimNames(sce) <- c("PCA_matrix", "UMAP")
reducedDimNames(sce)

# Calculate clusters from an SCE object using default parameters
cluster_results_df <- calculate_clusters(
  sce,
  pc_name = "PCA_matrix"
)
cluster_results_df
```

@@

Results can either be directly exported as a TSV file (e.g., with `readr::write_tsv()`), or you can add the results into your SCE or Seurat object.
The subsequent examples will show you how to save the `cluster_results_df` data frame to an SCE and a Seurat object.

@@

To avoid ambiguity between the existing and new clustering results, you may want to rename incoming columns.

We'll therefore begin by adding the prefix `ropenscpca_` to all column names in the `cluster_results_df` data frame before adding them to a given object.

@@

# Second, add columns as desired using the barcodes and cell_id columns to join
sce_coldata <- sce_coldata |>
  dplyr::left_join(
    cluster_results_renamed_df,
    by = c("barcodes" = "ropenscpca_cell_id")
  )

# Third, add the modified colData back into the SCE
# Use the row.names argument to ensure rownames (barcodes) are retained; don't forget this step!
colData(sce) <- DataFrame(
  sce_coldata,
  row.names = sce_coldata$barcodes
)

colData(sce)
```
